A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-11-19 paper/paper.tex
paper: kill some floats, adjust whitespace

  We exploit this correspondence to develop an original uniform indexed definition of both augmented semi-simplicial and semi-cubical sets, and fully formalize it in Coq.
\end{abstract}
\maketitle

@@

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
up to cubical faces identities. In it, $X_1$ can be seen as a family over $X_0 \times X_0$ presented in a fibered form, $X_2$ can be seen as a family of $X_1 \times X_1 \times X_1 \times X_1$, and so on. This suggests an alternative indexed presentation of the same presheaf as a stratified sequence of families indexed on elements of families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it takes the form:

\begin{align*}
  X_0 & :              & \U                                                                            \\
  X_1 & :              & X_0 \times X_0 \rightarrow \U                                                 \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \times X_1 (c)(d) \times X_1(a)(c) \times X_1 (b)(d) \rightarrow \U \\
  \ldots
\end{align*}

@@

\begin{definition}[$\DeltaPlus$]
  The definition of $\DeltaPlus$ is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj_{\DeltaPlus}       & := \Nat                                                                             \\
      \Hom_{\DeltaPlus}(p, n) & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
      g \circ f               & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                        \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases}                \\
      \id                     & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom_{\DeltaPlus}(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj_{\Cube}       & := \Nat                                                                          \\
      \Hom_{\Cube}(p, n) & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f          & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                               \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                           \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id                & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
  \end{figure}

@@

\begin{table}[H]
  \centering
  \begin{tabularx}{0.7\linewidth}{X|X|X}
    \toprule
    Cardinal of $\nu$ & 1                               & 2                  \\
    \graymidrule
    Interpretation    & Augmented semi-simplicial types & Semi-cubical types \\
    \bottomrule
  \end{tabularx}
\end{table}

@@

\begin{definition}[$\hexagon$]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj_{\hexagon}       & := \Nat                                                                           \\
      \Hom_{\hexagon}(p, n) & := \{l \in (\nu \cup \{\kstar\})^n \mid \text{number of $\kstar$ in $l = p$}\}    \\
      g \circ f             & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                                 \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a \in \nu$}                                \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a \in \nu$ or $a = \kstar$} \\
      \end{cases}   \\
      \id                   & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom_{\hexagon}(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

@@

In $\U[m]$, the following properties hold:

\begin{enumerate}
  \item UIP holds on the unit type, bool type, as well as all types of finite cardinal $\nu$.
  \item UIP propagates to $\Sigma$-types.
  \item UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on the building blocks which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$: they represent matching objects.

To every $X_n$ is uniformly assigned a type of the form $\fullframe[n] \rightarrow \U$. Compared to the informal description given in section~\ref{sec:rel-param}, this amounts to applying the isomorphisms between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A.\, (B a \rightarrow C)$ and $(\Sigma a: A.\, B a) \rightarrow C$ so as to get the expected form $\fullframe[n] \rightarrow \U$. In particular, $\fullframe[n]$ is a ``telescope'', i.e. a nesting of $\Sigma$-types.

Let us now illustrate how we recursively build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ given to $X_0$ in section~\ref{sec:rel-param} can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from the $\unittype$ type and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$, as shown below. The figure also mentions how the type of $X_2$ is structured.

\begin{figure}[H]
  \centering
  \small
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
\end{figure}

>>>>

  We exploit this correspondence to develop an original uniform indexed definition of both augmented semi-simplicial and semi-cubical sets, and fully formalize it in Coq.
\end{abstract}
\maketitle
\vspace{-2em} % Remove space after abstract

@@

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
up to cubical faces identities. In it, $X_1$ can be seen as a family over $X_0 \times X_0$ presented in a fibered form, $X_2$ can be seen as a family of $X_1 \times X_1 \times X_1 \times X_1$, and so on. This suggests an alternative indexed presentation of the same presheaf as a stratified sequence of families indexed on elements of families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it takes the form:
\begin{align*}
  X_0 & :              & \U                                                                            \\
  X_1 & :              & X_0 \times X_0 \rightarrow \U                                                 \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \times X_1 (c)(d) \times X_1(a)(c) \times X_1 (b)(d) \rightarrow \U \\
  \ldots
\end{align*}

@@

\begin{definition}[$\DeltaPlus$]
  The definition of $\DeltaPlus$ is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.
  \begin{align*}
    \Obj_{\DeltaPlus}       & := \Nat                                                                             \\
    \Hom_{\DeltaPlus}(p, n) & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
    g \circ f               & :=
    \begin{cases}
      f                  & \text{if $g = []$}                                                        \\
      0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
      a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
    \end{cases}                \\
    \id                     & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom_{\DeltaPlus}(n, n)$}
  \end{align*}
\end{definition}

@@

  \begin{align*}
    \Obj_{\Cube}       & := \Nat                                                                          \\
    \Hom_{\Cube}(p, n) & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
    g \circ f          & :=
    \begin{cases}
      f                  & \text{if $g = []$}                                                               \\
      a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                           \\
      a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
    \end{cases} \\
    \id                & := [\kstar, \ldots, \kstar] \text{ $n$ times}
  \end{align*}

@@

\begin{center}
  \begin{tabularx}{0.7\linewidth}{X|X|X}
    \toprule
    Cardinal of $\nu$ & 1                               & 2                  \\
    \graymidrule
    Interpretation    & Augmented semi-simplicial types & Semi-cubical types \\
    \bottomrule
  \end{tabularx}
\end{center}

@@

\begin{definition}[$\hexagon$]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.
  \begin{align*}
    \Obj_{\hexagon}       & := \Nat                                                                           \\
    \Hom_{\hexagon}(p, n) & := \{l \in (\nu \cup \{\kstar\})^n \mid \text{number of $\kstar$ in $l = p$}\}    \\
    g \circ f             & :=
    \begin{cases}
      f                  & \text{if $g = []$}                                                                 \\
      a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a \in \nu$}                                \\
      a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a \in \nu$ or $a = \kstar$} \\
    \end{cases}   \\
    \id                   & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom_{\hexagon}(n, n)$}
  \end{align*}
\end{definition}

@@

In $\U[m]$, the following properties hold:
\begin{enumerate}
  \item UIP holds on the unit type, bool type, as well as all types of finite cardinal $\nu$.
  \item UIP propagates to $\Sigma$-types.
  \item UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on the building blocks called $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$: they represent matching objects.

Every $X_n$ is uniformly assigned a type of the form $\fullframe[n] \rightarrow \U$. Compared to the informal description given in section~\ref{sec:rel-param}, this amounts to applying the isomorphisms between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a: A.\, (B a \rightarrow C)$ and $(\Sigma a: A.\, B a) \rightarrow C$ to get the expected form $\fullframe[n] \rightarrow \U$. In particular, $\fullframe[n]$ is a ``telescope'', or a nesting of $\Sigma$-types.

Let us now illustrate how we recursively build $\fullframe[n]$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ given to $X_0$ in section~\ref{sec:rel-param} can be equivalently formulated as $\unittype \rightarrow \U$. Then, more generally, we let each $\fullframe[n]$ consist of $n$ layers, written $\layer[n][p]$ with $p < n$, that we stack in order, starting from $\unittype$, and writing $\framep[n][p]$ for the $p$ first layers of a $\fullframe[n]$, so that $\fullframe[n]$ is $\framep[n][n]$. For instance, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of $\unittype$ and $\layer[1][0]$, as shown below. The figure also mentions how the type of $X_2$ is structured.
\begin{small}
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
\end{small}
